# GLMs for binary data 

## Binary data 
- Binary data is a type of binomial count data in which the binomial denominator is one 
  - Therefore, every trial has a value of either one or zero 
- This type of data can be analyzed using the GLM with a binomial distribution with the same set of link functions to prevent predictions below zero and above one 
- However, since this type of data is so constrained, there are some differences: 
  - The ratio of residual deviance: DF to diagnose under/overdispersion does not apply 
  - But we can use the `binnedplot()` function in the arm package to diagnose dispersion 

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
install.packages("arm",  repos = "https://cran.us.r-project.org")
install.packages("ggplot2",  repos = "https://cran.us.r-project.org")
install.packages("ggfortify",  repos = "https://cran.us.r-project.org")
install.packages("patchwork",  repos = "https://cran.us.r-project.org")
install.packages("dplyr",  repos = "https://cran.us.r-project.org")
install.packages("Sleuth3",  repos = "https://cran.us.r-project.org")
install.packages("readr",  repos = "https://cran.us.r-project.org")
```

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
library(arm)
library(ggplot2)
library(ggfortify)
library(dplyr)
library(patchwork)
library(Sleuth3)
library(readr)
```

## The wells data set for the binary GLM example 
- The example dataset for this chapter come from Gelman and Hill (2006): 
- The data concern a Bangladeshi region in whcih many drinking water wells are contaminated by naturally occuring arsenic: 
```{r}
urlfile="https://raw.githubusercontent.com/apicellap/data/main/Data_Binary_Wells.csv"
wells<-read_csv(url(urlfile))
str(wells)
```

- The dataframe consists of three important variables (the other two we will ignore):
  - switch - the binary response of whether or not people switch the well from which they were drinking 
  - arsenic - the concentration of arsenic 
  - dist - the distance to the nearest safe well 
  
Define some labels: 
```{r}
xlabel <- "Distance to nearest well"
ylabel <- "Probability of switching"
```

Visualize the data of how distance to a safe well may predict switching probability: 
```{r}
fig18_1 <- ggplot(wells, aes(x = dist, y = switch)) + 
  labs(x = xlabel, y = ylabel) + 
  geom_point() + 
  geom_smooth()
fig18_1
```

- From the figure it seems like distance to a well may influence whether someone switches 

Rescale the x axis values: 
```{r}
wells$dist100 <- wells$dist/100
```

Fit a GLM for switching vs distance: 
```{r}
fit_1 <- glm(switch ~ dist100, binomial(link = "logit"), data = wells)
```

Check the model's diagnostics: 
```{r}
autoplot(fit_1)
```

- Due to the constrained nature of the binary data, the diagnostic plots are of little value 

Diagnose dispersion: 
```{r}
x <- predict(fit_1) #extract predicted values from the model
y <- resid(fit_1) #extract the residuals from the model 
fig18_3 <- binnedplot(x, y)
```

- The gray lines on the plot indicate $\pm2$ standard errors within which approximately 95% of the binned residuals lie 
- Ideally for this type of graph, we'd want plenty of bins and plenty of datapoints per each bin, so this approach works well for smaller datasets 
- The expectation that around 95% of the residuals falls within the bounds seems to be met, so we can proceed: 

Examine the coefficents: 
```{r}
coef(fit_1)
```

Examine the 95% confidence interval: 
```{r}
confint(fit_1)
```

- This CI supports the hypothesis that the further away a well is, the less likely people will switch to it 
  - Gelman & Hill (2006) suggest a rule of thumb for interpretting the slope of a logistic regression: 
    - The 'divide by four rule' in which you divide the coefficient for the logistic regression slope by 4 to get an approximate estimate of the maximum predicted effect of a unit change in the predictor on the response 
    
    
Plot the data superimposed with a binomial GLM: 
```{r}
fig18_4 <- ggplot(wells, aes(x = dist, y = switch)) + 
    labs(x = xlabel, y = ylabel) + 
  geom_point() + 
  geom_smooth(method = "glm", method.args = "binomial")
fig18_4
```

Evaluate the effect of arsenic concentration: 
```{r}
fit_2 <- glm(switch ~ arsenic, binomial(link = "logit"), 
             data = wells)
display(fit_2)
```

- Given this model and the following figure, there also seems to be a clear positive effect of higher arsenic concentrations leading to greater probabilities of switching wells: 

```{r}
fig18_5 <- ggplot(wells, aes(x = arsenic, y = switch)) + 
    labs(x = "Arsenic concentration", y = ylabel) + 
  geom_point() + 
  geom_smooth(method = "glm", method.args = "binomial")
fig18_5
```

## Centering 
- Both the effect of arsenic concentration and distance to the nearest safe well can be examined in the same model 
- Prior to fitting the GLM, we can make life easier by centering the explanatory variables by subtracting their mean value 
  - This approach has advantages when a regression intercept value of zero is unhelpful or doesn't make sense 
  
```{r}
wells$c.dist100 <- wells$dist100 - mean(wells$dist100)
wells$c.arsenic <- wells$arsenic - mean(wells$arsenic)
head(wells)
```

- Considering arsenic and distance simultaneously introduces the possibility of an interaction:

```{r}
fit_5 <- glm(switch ~ c.dist100 + c.arsenic + c.dist100:c.arsenic,
       family = binomial, data = wells )
display(fit_5)
```

- `display(fit_5)` output:
  - second row - effect of a change of 100 m when arsenic is at the average level 
  - third row - effect of a unit change of arsenic for a well at average distance 
  - fourth row (interaction) - for every 100 m increase, there is an additional -0.18 for the arsenic coefficient 
- Compare ratios of estimates to their respective standard errors to get a sense of the effect of each coefficient 
  - The interaction term's estimate not even 2x as big as its SE, so while the interaction between contaminant concentration and distance is going in that direction, it probably isn't significant 
  - We can look at whether the null hypotheses can be rejected or not by checking the 95% CI: 

```{r}
confint(fit_5)
```




