# Conclusions

## Introduction 
- This text began with an analysis of the Space Shuttle *Challenger* distaster 
- The author concludes the book by re-analyzing the data using keypoints covered in the past 17 chapters 

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
install.packages("arm",  repos = "https://cran.us.r-project.org")
install.packages("ggplot2",  repos = "https://cran.us.r-project.org")
install.packages("faraway",  repos = "https://cran.us.r-project.org")
install.packages("patchwork",  repos = "https://cran.us.r-project.org")
install.packages("dplyr",  repos = "https://cran.us.r-project.org")
install.packages("Sleuth3",  repos = "https://cran.us.r-project.org")
```

```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
library(arm)
library(ggplot2)
library(faraway)
library(dplyr)
library(patchwork)
library(Sleuth3)
```

## A binomial GLM analysis of the *Challenger* binary data
- The fuel leaks key to this dataset can be expressed in binary form

```{r}
urlfile="https://raw.githubusercontent.com/apicellap/data/main/ex2011.csv"
ex2011x<-read.csv(url(urlfile))
str(ex2011x)
```

```{r}
head(ex2011x)
```

This dataset contains an error, rectify it: 
```{r}
ex2011x[4,2] <- "Yes" #It is always better document changes in a script and not alter the original dataframe
```

Plot Failure vs. launch temperature: 
```{r f1, fig.cap=""}
fig19_1 <- ggplot(ex2011x, aes(x = Temperature, y = Failure)) + geom_point()
fig19_1
```

Recharacterize binary form of data: 
```{r}
ex2011x <- ex2011x %>% 
  mutate(failure = if_else(ex2011x$Failure == "Yes",1 , 0)) #if this statement is true then in column entitled 'failure' put 1, else put 0 
head(ex2011x)
```

Create a binomial GLM: 
```{r}
m1 <- glm(failure ~ Temperature, family = binomial(link = "logit"), data = ex2011x)
```

Extract the model's coefficients: 
```{r}
coef(m1)
```

Examine the 95% CI: 
```{r}
confint(m1)
```

- Based on the negative slope and the 95% CI, there is clearly a relationship between failure and temperature 
  - The probability of a fuel leak increases in colder conditions 
  
Plot the relationship: 
```{r fig.cap=""}
fig19_2 <- ggplot(ex2011x, aes(x = Temperature, y = failure)) + geom_point() + 
  geom_smooth(method = "glm", method.args = list(family = "binomial"))
fig19_2
```

- The plot shows a wide degree of uncertainty around the regression line 
- Even though there is a lot of uncertainty, the uncertainty encompasses a high degree for the system to fail 
- The temperature at the day of catastrophic launch was 30 degrees, which is well below the scope of the model 
- Still we can extrapolate from the data (taking with it a large grain of salt) through the binomial GLM 

Taking sample size into account we can get the t value for a two-tailed 95% CI when n = 21: 
```{r}
t21 <- qt(0.975, 21)
t21
```

Use the `predict()` function to make predictions from the GLM: 
```{r}
predicts <- predict(m1, data.frame(Temperature = 30:85), 
                    se = TRUE)
```

Create a dataframe of predicted ranges of values around the regression curve 
```{r}
fit <- ilogit(predicts$fit) #ilogit() back transforms from logit to probability values 
upper <- ilogit(predicts$fit + t21 * predicts$se.fit)
lower <-ilogit(predicts$fit - t21 * predicts$se.fit)
predictions <- data.frame(fit, upper, lower, Temperature = 30:85)
head(predictions)
```

Redraw the figure with extrapolated probability values for the new range of Temperature values: 
```{r f3, fig.cap=""}
fig19_3 <- ggplot(ex2011x, aes(x = Temperature, y = failure)) + 
  xlim(30, 80) + 
  geom_point() + 
  geom_line(data = predictions, aes(Temperature, fit), colour = "blue") + 
  geom_line(data = predictions, aes(Temperature, upper), colour = "darkgrey") + 
  geom_line(data = predictions, aes(Temperature, lower), colour = "darkgrey") 
fig19_3

```

- The prediction even at this lowest level of confidence is an unacceptably high probability of failure

Check assumptions: 
```{r fig.cap=""}
x <- predict(m1)
y <- resid(m1) 
fig19_4 <- binnedplot(x,y)
```

- Because most of the values lie outside of the binned zones, the binomial GLM is not a good model for this data (even though they are binary) 
- Other attempts have been made to model the data (Dalal et al. 1989), but there isn't yet an ideal solution for this data 
- Still, even with residuals not supporting a model, it was a poor decision to not even look at any of the data and recommend a launch at on day with much lower launch temperature than ever tested before 
- The author goes on to talk about how statistics is just one tool that must be combined with others 
  - For example, these o-rings placed in ice water become brittle too 
  - Other disciplines could have been consulted too 
- An imperfect statistical analysis could have saved the lives of seven astronauts 

## Recommendations 

1. *Make your research reproducible*
  - This can be accomplished through the use of Rmarkdown/quarto to make research reports 
2. *A picture is worth a thousands words* 
  - Graphs are the best way to present data 
  - Graph data before analyzing it to understand it 
  - Use graphs to assess assumptions/test diagnostics 
3. *Keep it simple*
  - Take the time to explore the data and look at estimates/intervals 
  - Get a sense of effect sizes and uncertainties 
4. *Consider more than one model*
  - Present 2-3 of the best models to help showcase the uncertain path towards model selection
5. *Attempt the P-free challenger*
  - Focus more on effect sizes and estimates 
6. *Report estimates, intervals, and sample sizes* 
  - The reproducibility crisis could be combatted by the reporting of these elements in the results section 
    - This would facilitate meta-analysis 
7. *Go back to basics* 
  - Understand the basics of error bars and intervals plus how to interpret them 
8. *Make a focused plan of analysis*
  - Plan how data will be analyzed in advance 
  - Meet all assumptions 
9. *Give P-values the respect they deserve (and no more)* 
  - If assumptions are not met, you end up with mushy p-values 
  - The use of p-values should be implemented when all assumptions are met and the satisfaction of all conditions that they depend on so that the values are as meaningful as possible 
10. *Focus on repeatability, not small p-values* 
  - No matter how small a p-value (or how high the level of confidence), the result can still be a false positive 
  - Scientific results are truly established when it is shown that they are repeatable
  - Author says that science values novelty too much at the expense of establishing a solid foundation 

