# Reproducible Research 

## The reproducibility crisis
- A concerning amount of research studies have been discovered over the last decade to not be reproducible 
- One aim of this chapter is to demonstrate how to write a script and then convert it into an Rmarkdown document 

```{r installation-04, echo=T, include=FALSE, message=FALSE, warning=FALSE, results='hide'}
install.packages("ggplot2",repos = "https://cran.us.r-project.org")
install.packages("SMPracticals",repos = "https://cran.us.r-project.org")
```

```{r load-libs-04, include=FALSE, echo=T, message=FALSE, warning=FALSE, results='hide'}
library(ggplot2)
library(SMPracticals)
```

## R Scripts 

Summary statistics on dataframe about atmospheric carbon dioxide concentrations: 
```{r summstats-04}
summary(co2)
```

Plot the data: 
```{r baseplot-04}
plot(co2)
```

## Analysis notebooks 
- The author says that one way to combat the reproducibility crisis is to document exactly how data analysis is done
- This can be accomplished through RMarkdown 
- Scripts can still be written in .R files, but should be converted to .Rmd files too

## RMarkdown
- He gives a basic overview of RMarkdown including how to convert it into a pdf
- He emphasizes that R is updated roughly every six months

Record which version of R was used to create script running: 
```{r sessinfo-04}
sessionInfo()
```


